---
title: "HW2-EDA with R - Part1: Plotting"
author: "Nivedita Chawdhari"
date: "February 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment
###[Kaggle Data Set](https://www.kaggle.com/muonneutrino/us-census-demographic-data/data)

## Step 2 - Create a new R Markdown document
RMD File : **HW2_EDAwR_Chawdhari.Rmd**

Working Directory : **HW2_files**

**[GitHub URL](https://github.com/nivi1990/HW2_Chawdhari)**


## Step 3 - Set working directory and explore data folder

 Data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data.



## Step 4 - Complete the following R tasks and answer questions

Let's load a few libraries we'll need:

```{r}
library(dplyr) 
library(ggplot2)
library(reshape2)
library(ggcorrplot)

```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
 county_df <- read.csv("data/county.csv")
county_MI <- read.csv("data/county_MI.csv")
```

Are there any NA values? What differences do you notice between the structure of the two data frames?

**There are no NA's in the countyMI data set while there are three NA's for the Income and IncomeError and Poverty Field each.**
**Structure wise there is an extra Peninsula column in the CountyMI data set,as a result Region Column does not have any categories**

###**Structure and Summary Check for the Dataframes:**
```{r}
str(county_df)
str(county_MI)

summary(county_df)
summary(county_MI)

#Replacing NA's with Mean 
county_df$Income <- ifelse(is.na(county_df$Income), mean(county_df$Income, na.rm=TRUE), county_df$Income)
county_df$IncomeErr <- ifelse(is.na(county_df$IncomeErr), mean(county_df$IncomeErr, na.rm=TRUE), county_df$IncomeErr)
county_df$ChildPoverty <- ifelse(is.na(county_df$ChildPoverty), mean(county_df$ChildPoverty, na.rm=TRUE), county_df$ChildPoverty)

```


### Problem 2 - Basic plots for Income

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US.Write a short paragraph describing the distribution of median incomes.

###Distribution of Mean Income
- **50 % of the people have median household income from 38826.25 to 52249.25**
- **There are many outliers on the higher side i.e people earning more than 75000**
- **The histogram is skewed to the right.**
- **There is an income inequality,this can be seen from the box plot,range for 4 th quartile  is way greater than combined range for first 3**      **quartiles i.e 75% of the people had a household income that falls between a short range while household income for rest 25% is spread out**


```{r}
#Histogram
ggplot(data=county_df, aes(county_df$Income)) + 
  geom_histogram(fill="Steel Blue" ,color="Black")+
  labs(title="Histogram for Median Income for US" , x="Income")

#Box Plot
ggplot(data =  county_df, aes(x = "US",y = county_df$Income)) +
  geom_boxplot(colour="black",fill="light blue",outlier.colour = "red")+
  labs(title = "Income Distribution", y="Income" )


#Denisity Plot
ggplot(data=county_df, aes(county_df$Income)) + 
  geom_density(fill="Steel Blue",alpha=0.8)+
  labs(title="Histogram for Median Income for US" , x="Income")


#Violin Plot
ggplot(data=county_df, aes(y = county_df$Income,x = "US")) + 
  geom_violin(fill="Light Blue") +
    labs(title="Violin Chart for Median Household Income for US" , y="Income")

quantile(county_df$Income)


```
  
### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US.

```{r histo_totpop}
ggplot(data=county_df, aes(county_df$TotalPop)) + 
  geom_histogram()+
  labs(title="Population by Counties in US",x="Total Population",y="Number of Counties")


```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
ggplot(data=county_df, aes(county_df$TotalPop)) + 
  geom_histogram(colour ="Red", fill ="light blue") +
  labs(title="Distribution of population for US counties" ,x="Total Population",y="Number of counties")

```

**Hacker Extra** - 

**kernel density plot overlaid on a histogram of Income**
```{r histo_OverlaidDensity}
 ggplot(county_df,aes(x=log10(county_df$TotalPop) )) +
  geom_histogram(aes(y=..density..),fill ="Steel Blue",color ="Gray")+
geom_density(color="Red",fill="Red",alpha=0.2) +
  labs(title="Density Overlaid Histogram for Total Population" ,x="Total Population")


```


Repeat the histogram but only for counties with less than 2.5 million people.

```{r histo_lt2.5m}

ggplot(county_df[county_df$TotalPop < 2500000,],aes(x=TotalPop)) +
  geom_histogram(colour="red", fill ="Light Blue")+
  labs(title="Distribution of population for US counties less than 2.5 Million" ,x="Total Population",y="Number of counties")
 

```


### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped box plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?

**Yes ,the peninsulas do differe with respect to the median income:**

- **The range of income for lower peninsula is greater than that for upper peninsula**
- **Overal Income for Lower Peninsula is higher than that for Upper Peninsula**
- **There are a few extreme income values for lower peninusla while for Upper Peninsula there are no outliers**
- **There is more income inequality in Lower Peninsula as compared to Upper Peninsula,The Income for Upper Peninsula seems to uniformly distributed as compared to LP**


```{r bp_inc_by_peninsula}
ggplot(data =  county_MI, aes(x = county_MI$peninsula,y = county_MI$Income)) +
  geom_boxplot(colour="black",fill="light blue",outlier.colour = "red")+
  labs(title = "Michigan Income by Peninsula",x="Peninsula" , y="Income in Million" )

```

Create several more boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

- **Lower Peninusla has more counties with higher population as compared to UP,this can be due to the presence of Urban and Industrial area in the LP**
- **The mean Commute time is greater overallfor LP than the UP,this can be due to more number bigger cities in LP more spread out     industrail area**
- **The percentage of population that is Hispanic is greater in the LP whereas the Native American population is greater in UP**
- **Percentage of Private work is less in LP due the absence of Industrial area and hence the Public Work % is greater in UP as compared to LP**
- **It can be seen that for both peninsulas Driving is primary means of commute followed by Car Pooling**


```{r more_boxplots,warning=FALSE}
#Boxplot for Total Population
ggplot(data =  county_MI, aes(x = county_MI$peninsula,y = county_MI$TotalPop)) +
  geom_boxplot(colour="black",fill="light blue",outlier.colour = "red")+
  scale_y_continuous(limits = c(0, 250000))+
  labs(title = "Michigan Population by Peninsula",x="Peninsula" , y="Total Population" )

#Boxplot for Mean Commute
ggplot(data =  county_MI, aes(x = county_MI$peninsula,y = county_MI$MeanCommute)) +
  geom_boxplot(colour="black",fill="light blue",outlier.colour = "red")+
  labs(title = "Michigan Mean Commute Peninsula",x="Peninsula" , y="Mean Commute" )

#Boxplot for Public Work
ggplot(data =  county_MI, aes(x = county_MI$peninsula,y = county_MI$PublicWork)) +
  geom_boxplot(colour="black",fill="light blue",outlier.colour = "red")+
  labs(title = "Michigan Public Work Percentage by Peninsula",x="Peninsula" , y="Public Work" )



#Grouped plot for Population Race
mdata <- melt(data = county_MI, id = 'peninsula', measure = c ('Hispanic','White','Black','Native','Asian'))
 ggplot(mdata) + geom_boxplot(aes(x=peninsula, y= log(value), color=variable)) +
  labs(title = "Distribution of Races By Peninsula" , y="log(% of Race)" )
 
 
 #Grouped Plot for Means of Commute
 mdata <- melt(data = county_MI, id = 'peninsula', measure = c ('Drive','Carpool','Transit','Walk','OtherTransp'))
 ggplot(mdata) + geom_boxplot(aes(x=peninsula, y=log(value), color=variable)) +
   labs(title = "Distribution of type of Commute By Peninsula" , y="log(% of Type of Commute Used)" )
  
 
 
  
 
 
```

### Problem 5 - How do regions of the country compare?

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how some variable of interest to you compares by region. For 
example, you could explore how the regions compare with respect to the
distribution of MeanCommute time. Make sure your graph has a title and good axis
labels.


**For all the regions data seems to be distributed similarly for most of the variables of interest**
**For a few variables like number of white population the data is skewed left. South has  more concentrated white population than other region.**
**Overall poverty is between 20 to 40 % in all four regions**


```{r faceted_histo}
ggplot(data=county_df, aes(county_df$Employed) ) + 
  geom_histogram(colour ="Red", fill ="light blue") + 
  labs(title="Distribution of Employment for US counties by Region" ,x="Employed",y="Number of counties") + 
  theme(axis.text.x = element_text(angle=90))+
  facet_grid(. ~ Region)

ggplot(data=county_df, aes(county_df$Poverty)) + 
  geom_histogram(colour ="Red", fill ="light blue") + 
  labs(title="Distribution of Poverty ratio for US counties by Region" ,x="Poverty",y="Number of counties") + 
  facet_grid(. ~ Region)

ggplot(data=county_df, aes(county_df$White)) + 
  geom_histogram(colour ="Red", fill ="light blue") + 
  labs(title="Distribution of White Race population for US counties" ,x="White",y="Number of counties") + 
  facet_grid(. ~ Region)


ggplot(data=county_df, aes(county_df$MeanCommute)) + 
  geom_histogram(colour ="Red", fill ="light blue") + 
  labs(title="Distribution of Mean Commute Time for US counties" ,x="MeanCommute",y="Number of counties") + 
  facet_grid(. ~ Region)

ggplot(data=county_df, aes(county_df$IncomePerCap)) + 
  geom_histogram(colour ="Red", fill ="light blue") + 
  labs(title="Distribution of Income Per Capita for US counties" ,x="IncomePerCapita",y="Number of counties") + 
  facet_grid(. ~ Region)



```


### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Region in the US. Make sure
your graph has a title and good axis labels.

```{r bar_counties_by_region}
ggplot(county_df,aes(x = county_df$Region)) + 
  geom_bar(fill="steel blue") + 
  labs(title = "Number of Counties By Region", x="Region" , y="Number of Counties") +
  theme_minimal()

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(county_MI,aes( y =county_MI$Income,x=county_MI$MeanCommute)) + 
  geom_point() + 
  labs(title = "Scatter plot for Mean Commute and Income", x="MeanCommute" , y="Income") +
  theme_minimal()+
  geom_smooth()

cor(county_MI$Income,county_MI$MeanCommute)


```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula.

###Relationship between Income and MeanCommute
**The  correlation relation is not strong enough,but a positive trend can be noticed for the variables,**
**for UP The Income range is lower with less commute time,while for LP the income range is higher with greater commute time as compared to UP.**


```{r xy_inc_commute_peninsula}

ggplot(county_MI,aes(color=county_MI$peninsula,y = county_MI$Income,x= county_MI$MeanCommute)) + 
  geom_point() + 
  labs(title = "Mean Commute vs Income for Michigan", x="MeanCommute" , y="Income") +
  geom_smooth()+
  theme_minimal()

```

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

**A strong trend can be observed via the scatter plot for these two pairs:**

**1. Public Work and Employed : these variables seems to have a negative relation for both the peninsulas,counties with higher public work percentage has lower employment rate,this relation is stronger for the LP than the UP**

**2.Asian and Employed :  variables seems to have a positve relation for both the peninsulas,counties with higher percentage of Asian population has a higher employment rate,this relation is stronger for the LP than the UP,in this case too the relation is stronger for LP than the UP**


```{r xy_faceted_peninsula,warning=FALSE}


ggplot(county_MI,aes(x=  log10(county_MI$PublicWork) ,y =log10(county_MI$Employed) )) + 
  geom_point() + 
  labs(title = "Relation between Public work and Employment rate", x="PublicWork" , y="Employed") +
  geom_smooth()+
  facet_grid(~ county_MI$peninsula)


ggplot(county_MI,aes(x= (county_MI$Asian),y = county_MI$Employed  )) + 
  geom_point() + 
  labs(title = "Relation Between Asian race and Emloyment", x="Asian" , y="Employed") +
  geom_smooth()+
    xlim(0, 4)+
  facet_grid(~ county_MI$peninsula)


```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
#Wihout Log scale 
ggplot(county_df,aes( y = (county_df$Income),x= (county_df$TotalPop) )) + 
  geom_point() + 
  labs(title = "Scatter plot for Total Population and Income", x="TotalPopulation" , y="Income") +
  theme_minimal()+
  geom_smooth()
cor(county_df$TotalPop,county_df$Income)


#With Log Scale 
ggplot(county_df,aes( y = log10(county_df$TotalPop),x= log10(county_df$Income) )) + 
  geom_point() + 
  labs(title = "Scatter plot for Total Population and Income", x="Income" , y="Total Population") +
  theme_minimal()+
  geom_smooth()
cor(log10(county_df$TotalPop),log10(county_df$Income))

```
###Relation between Total Popluation and Income 
**The total population and Income do not seem to show any strong relation, although plotting the data makes it improved the pattern for the scatter,and it shows somewhat positive trend.**
**The correlation coeffienct also shows no stronger relation between the two variable**

###Why Log Scale??
- **There are a few data points those have very large numbers for total population that differs greatly from rest to the data points.**
- **These outlier makes the graph look skewed to the right,also the compressed data points makes it difficult to notice small changes and bias the    the judgement of the pattern.**
- **The difference between the outliers and other data points is huge,this difference will hence domiate the difference between the other data -    points when plotted on a normal scale.**
- **Log scale on the other hand plots the ratio rather than the difference ,it normalized the exponential data.**
- **Since the data for Income is in orders of magnitude,using log scale wil help getting the whole data on one axis.**
- **The log scale improves the correlation coeffienct for the variables,it also improves the pattern.**

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.

**I did a Correlation Matrix of a few variables,out of which Income,Professionals has a positive correlation.**
**Also another pairs of interest for Michigan data me were Type of Work (Public,Private) with % that use Walking as means of Commute,there is a positive** **Correlation for Public work and Walk while Negative for Private and Walk ,this suggest that people working in Private sector domiated the** **drive percentage of commute method while those working in Public sector use walking as means of commute**
**Correlation Matrix**
```{r}
Cordata <-county_df[,c(7,17,23:28,31,34,36,37)]
CorDatMI<- county_MI[,c(8,18,24:29,32,35,37,38)]
cor(Cordata)
cor(CorDatMI)

```

**The Relation between Total Population to Income looks better for Michigan only data as compared to the US data set**
```{r}
ggplot(county_MI,aes( y = log10(county_MI$TotalPop),x= log10(county_MI$Income) )) + 
  geom_point() + 
  labs(title = "Scatter plot for Total Population and Income", x="Income" , y="Total Population") +
  theme_minimal()+
  geom_smooth()


```



